---
title: "Assignment 1"
author: "Arkaprabha Gun"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document: 
    latex_engine: xelatex
    keep_tex: true
    extra_dependencies: ["booktabs", "longtable", "array", "multirow", "wrapfig", "float", "colortbl", "pdflscape", "tabu", "threeparttable", "threeparttablex", "fontspec", "makecell", "xcolor"]
  html_document: default
---

```{r setup, include=FALSE}
if (!require("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

pacman::p_load(ggdag, dagitty, ggplot2, tidyverse, wesanderson,
               epiR, broom, EValue, stdReg, gtsummary, broom.helpers,epiR,haven,survival,survminer,kableExtra)
addicts2 <- read_dta("data/addicts2.dta")

summary(addicts2)
options(tinytex.verbose = TRUE)
hist(addicts2$id,breaks = unique(addicts2$id),freq = TRUE)
set.seed(123)
```


# Question1 Generate Kaplan-Meier curves for each clinic to visualize the data.

### Question 1 a Show the graph, including confidence intervals.

```{r}
addicts2[addicts2$clinic %in% c(9, 22), ]
addicts2 <- addicts2[addicts2$clinic %in% c(1, 2) & addicts2$survt > 0 & addicts2$survt < 2000, ]


addicts_clean <- addicts2[addicts2$clinic %in% c(1, 2), ]

km_fit <- survfit(Surv(survt, status) ~ clinic, data = addicts_clean)

km_fit

ggsurvplot(
  km_fit,
  data = addicts_clean,
  conf.int = TRUE,           
  pval = FALSE,              
  risk.table = FALSE,        
  xlab = "Time (days)",
  ylab = "Survival Probability",
  title = "Kaplan-Meier Survival Curves by Clinic",
  legend.title = "Clinic",
  legend.labs = c("Clinic 1", "Clinic 2")
)
```

## Question 1b) Perform a log-rank test to formally evaluate whether the two curves are statistically different. What is your conclusion?

```{r}
logrank_test <- survdiff(Surv(survt, status) ~ clinic, data = addicts_clean)

logrank_test
```

The two survival curves are statistically significantly different (p \< 0.001). We reject the null hypothesis that the Kaplan-Meier(survival) curves are the same between the two clinics.

## Question 2) Generate a graph of the conditional hazards. Show the graph. Do the hazards appear proportional? (/6)

```{r}
plot(km_fit, fun="cumhaz", col=c("red", "blue"), lwd=2,
     xlab="Time (days)", 
     ylab="Cumulative Hazard",
     main="Cumulative Hazard Functions by Clinic")
legend("topleft", c("Clinic 1", "Clinic 2"), col=c("red", "blue"), lty=1, lwd=2)
```

The hazards do not appear to be proportional, Since the curves are not parallel. Hazards that are proportional would be represented by curves between which the angles remain constant.

## Question 3) Plot ln[-ln S(t)] against ln(t).

### a) Show the graph.

```{r}
plot(km_fit, fun="cloglog", col=c("red", "blue"), lwd=2,
     xlab="ln(Time)", 
     ylab="ln(-ln(Survival))",
     main="Complementary Log-Log Plot by Clinic")
legend("topleft", c("Clinic 1", "Clinic 2"), col=c("red", "blue"), lty=1, lwd=2)
```

### b Which of the Weibull, exponential, proportional hazards and/or accelerated failure-time assumptions hold? Please explain your reasoning for each of these assumptions. (/3)

Proportional Hazards (PH) assumption does not hold. For PH to hold, the lines must be parallel. The lines are not parallel, the converge and then diverge again.This confirms what we saw in the cumulative hazard plot.
Similarly, the Exponential distribution assumption For an exponential distribution, the lines would need to be straight, parallel, and horizontal (zero slope). These lines are curved, not parallel, and have positive slopes.
The Weibull distribution assumption does not hold. For a Weibull distribution, the lines should be straight and parallel (though they can have a slope). These lines show curvature and are not parallel.Accelerated Failure Time (AFT) assumption dpes not hold AFT models require parallel lines in this plot (similar to PH). Since the lines are not parallel, the AFT assumption does not hold either.

### c What types of models can you use with these data? (/2)

The choice of models we could potentially use are ones that allow the proportional hazards assumption to be violated. Some examples would be a cox-proportional hazard model with a time trend for the hazard, or parametric models like splines that allow us to fit without assuming a specific distribution ex-ante.AFT models that do not require proportional hazards by virtue of their model specification could also be potentially used.

## Question 4 Run a Cox model adjusting for prison and methadone dose.

### a Show your results in a table format. (/4)

```{r}
cox_model <- coxph(Surv(survt, status) ~ clinic + prison + dose, data = addicts2)

tbl_regression(cox_model, 
               exponentiate = TRUE,
               label = list(
                 clinic ~ "Clinic (2 vs 1)",
                 prison ~ "Prison Record (Yes vs No)",
                 dose ~ "Methadone Dose (per mg/day)"
               ))%>%
  as_kable_extra(booktabs = TRUE, longtable = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

### b Formally test the proportional hazards assumption. Show your results. Does the assumption hold? If not, for which variables does the assumption not hold? (/6)

```{r}
ph_test <- cox.zph(cox_model)

print(ph_test)

ggcoxzph(ph_test)
```

The proportional hazards assumption does NOT hold for the overall model (p = 0.005). Specifically, the assumption is violated for the clinic variable (p \< 0.001), while it holds for prison and dose.

### c If the assumption does not hold, run another model taking this into account. Show your results (hazard ratios and 95% confidence intervals) in a table for all 3 variables (clinic, prison and methadone dose) and interpret your results in a few sentences. Which of clinic 1 or 2 appears to perform better? (/8)

```{r}
cox_tv <- coxph(Surv(survt, status) ~  clinic +prison + dose  + tt(clinic), 
                data = addicts2,
                tt = function(x, t, ...) x * log(t))

tbl_regression(cox_tv, 
               exponentiate = TRUE,
               label = list(
                 prison ~ "Prison Record (Yes vs No)",
                 dose ~ "Methadone Dose (per mg/day)",
                 clinic ~ "Clinic 2 vs 1 (baseline)",
                 `tt(clinic)` ~ "Clinic 2 vs 1 × log(time) (time-varying)"
               ))%>%
  as_kable_extra(booktabs = TRUE, longtable = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

At baseline (t=1 day), patients in Clinic 2 have a 661% higher hazard of dropping out compared to Clinic 1  (HR = 7.61, 95% CI: 0.98-59.3, p = 0.053). However, this initial difference in hazard changes over time, as indicated by the significant time-varying component (HR = 0.58, 95% CI: 0.40-0.84,). Patients with a prison record have a 45% higher hazard of dropping out (HR = 1.47, 95% CI: 1.05-2.04). Each additional mg/day of methadone dose is associated with a 3% reduction in dropout hazard (HR = 0.97, 95% CI: 0.96-0.98, p \< 0.001). We can conclude that while the Hazard in Clinic 2 is higher initially, it performs better in ensuring survival over a period of longer follow-up.

## Question 5 Run a hazard model assuming an exponential distribution. Hint: use flexsurvreg

### a Show your results (estimate and 95% confidence intervals) in a table for all 3 variables (clinic, prison and methadone use) and interpret your results in a few sentences. Which of clinic 1 or 2 appears to perform better? (/8)9i

```{r}
p_load(flexsurv)

exp_model <- flexsurvreg(Surv(survt, status) ~ clinic + prison + dose, 
                         data = addicts_clean, 
                         dist = "exponential")



exp_results <- data.frame(
  Variable = c("Clinic (2 vs 1)", "Prison (Yes vs No)", "Dose (per mg/day)"),
  Estimate = coef(exp_model)[c("clinic", "prison", "dose")],
  HR = exp(coef(exp_model)[c("clinic", "prison", "dose")])
)

ci <- confint(exp_model)
exp_results$Lower_CI <- exp(ci[c("clinic", "prison", "dose"), 1])
exp_results$Upper_CI <- exp(ci[c("clinic", "prison", "dose"), 2])

exp_results
```

Clinic 2 performs significantly better than Clinic 1. Patients in Clinic 2 have 58% lower hazard of dropping out compared to Clinic 1 (HR = 0.420, 95% CI: 0.278-0.635). This indicates substantially better patient retention in Clinic 2. Patients with a prison record have 30% higher hazard of dropping out, though this difference is not statistically significant (HR = 1.305, 95% CI: 0.943-1.805). Each additional mg/day of methadone dose reduces the dropout hazard by 2.8% (HR = 0.972, 95% CI: 0.960-0.983).

## 5 b Graph the predicted hazard by time using the plot command. Show the graph. (/4)

```{r}
plot(exp_model, type="hazard", ci=TRUE)

```

### 5c) As a reminder, the exponential regression model assumes that the baseline hazard (the hazard when all covariates equal zero) is constant over time.  Based on the results that you have obtained so far, does the evidence support that assumption or not?  Justify your answer in one or two sentences. (/4)

The evidence does not support the exponential assumption. The complementary log-log plots showed non-linear patterns and non-parallel curves, indicating the hazard is not constant over time. Additionally, the cumulative hazard functions diverged rather than remaining linear and parallel as required by the exponential distribution.

## Question 6 Run an accelerated failure-time (AFT) model assuming a Weibull distribution.

### 6 a Show your results (estimate and 95% confidence intervals) in a table for all 3 variables (clinic, prison and methadone use) and interpret your results in a few sentences. Which of clinic 1 or 2 appears to perform better? (/8)

```{r}
weibull_aft <- flexsurvreg(Surv(survt, status) ~ clinic + prison + dose, 
                           data = addicts2, 
                           dist = "weibull")

library(broom)
library(dplyr)

weibull_results <- tidy(weibull_aft, conf.int = TRUE)

weibull_cov <- weibull_results %>%
  mutate(
    exp_estimate = exp(estimate),
    exp_lower = exp(conf.low),
    exp_upper = exp(conf.high),
    Variable = case_when(
      term == "shape" ~ "Shape Parameter",
      term == "scale" ~ "Scale Parameter",
      term == "clinic" ~ "Clinic (2 vs 1)",
      term == "prison" ~ "Prison Record (Yes vs No)",
      term == "dose" ~ "Methadone Dose (per mg/day)",
      TRUE ~ term
    )
  ) %>%
  select(Variable, estimate, exp_estimate, conf.low, conf.high, 
         exp_lower, exp_upper, p.value)

kable(weibull_cov, 
      format = "latex",
      booktabs = TRUE,
      caption = "Weibull AFT Model Results") %>%
  kable_styling(latex_options = "hold_position")
```

The shape parameter of 1.38 (95% CI: 1.20-1.57) indicates the baseline hazard increases over time, suggesting dropout risk accelerates the longer patients remain in treatment. Patients in Clinic 2 have survival times that are approximately twice as long as those in Clinic 1 (acceleration factor = 2.01, 95% CI: 1.48-2.72, p \< 0.001). Patients with a prison record have survival times that are 21% shorter (acceleration factor =  0.79, 95% CI: 0.62-1.00, p = 0.048), indicating they drop out faster. Each additional mg/day of methadone dose increases survival time by 2.5% (acceleration factor = exp(0.02) = 1.025, 95% CI: 1.015-1.033).

### 6b Graph the predicted hazard by time using the plot command. Show the graph. Compare this graph with the one you would expect for the exponential model. Are they similar or different? Why do you think that is?

```{r}
plot(weibull_aft, type="hazard", ci=TRUE)

```

The graph from the exponential model shows a constant hazard over time (which we have assumed). Whereas, the Weibull model has a shape parameter = 1.38(95% CI: 1.20-1.57) . When shape \> 1, the hazard increases over time. The exponential is a special case of Weibull where shape = 1 (constant hazard). Since our data has shape = 1.38, the hazard increases over time, making the Weibull a better fit than the exponential.

### 6 c Given what you know about the mathematical forms of the Weibull and exponential models, what information can allow you to test whether the baseline hazard is best approximated by a Weibull than an exponential distribution? Hint: look at residuals from the Weibull model.

```{r}

library(broom)
weibull_results <- tidy(weibull_aft, conf.int = TRUE)
weibull_results %>% filter(term == "shape")


AIC(exp_model)      
AIC(weibull_aft)    


cs_resid_weibull <- residuals(weibull_aft, type = "response")
cs_resid_exp <- residuals(exp_model, type = "response")


km_resid_weibull <- survfit(Surv(cs_resid_weibull, addicts2$status) ~ 1)
km_resid_exp <- survfit(Surv(cs_resid_exp, addicts2$status) ~ 1)

par(mfrow = c(1, 2))

plot(km_resid_weibull$time, -log(km_resid_weibull$surv), 
     type = "s", xlab = "Cox-Snell Residuals", 
     ylab = "Cumulative Hazard",
     main = "Weibull Model Residuals")
abline(0, 1, col = "red", lty = 2)

plot(km_resid_exp$time, -log(km_resid_exp$surv), 
     type = "s", xlab = "Cox-Snell Residuals", 
     ylab = "Cumulative Hazard",
     main = "Exponential Model Residuals")
abline(0, 1, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

Yes, the baseline hazard is significantly better approximated by the Weibull than the exponential distribution. The shape parameter confidence interval excludes 1.0, and the Weibull model has substantially better fit (lower AIC, 2151.67 vs 2168.95).When we plot the cox-snell residuals, we see that both of them fit similarly well, and are difficult to differentiate visually. But on the basis of the other evidence, we can find that the bazeline hazard is better modelled using a weibull model than an exponential one.

### 6d Weibull models can be used to model survival and hazard rates. To model hazard rates, run a proportional hazards model using Weibull AFT. You will notice that the sign (positive vs negative) of the coefficients for the independent variables are all reversed in the hazard form relative to the AFT form that you obtained above. Why do you think that is?

```{r}
weibull_ph <- flexsurvreg(Surv(survt, status) ~ clinic + prison + dose, 
                          data = addicts_clean, 
                          dist = "weibullPH")

tidy(weibull_ph, exponentiate = TRUE, conf.int = TRUE)

library(broom)
aft_coef <- tidy(weibull_aft)[c(3:5), c("term", "estimate")]
ph_coef <- tidy(weibull_ph)[c(3:5), c("term", "estimate")]

comparison <- merge(aft_coef, ph_coef, by = "term", suffixes = c("_AFT", "_PH"))
comparison
```

AFT models model the effect of a particular exposure/intervention on survival time. Whereas hazard models model the effect on hazard, which is the risk of failure. Hence the two concepts are inversely related/are complements to each other. As a result we see that the sign of the estimated coefficient are reversed in the AFT model compared to the hazard models, even though they are modelling the same relationship.

# Section II Survival Curves

```{r}
 library(haven)
divorce_EPI608 <- read_dta("data/divorce_EPI608.dta")


summary(divorce_EPI608)

length(unique(divorce_EPI608$id))

```

## Analysis 
### Step 1	Run a pooled logistic regression model that estimates the conditional probability of remaining married including all covariates, a time variable and an interaction term between education and time to allow for time-varying HR.  If you can include a flexible functional form for time (such as a spline), great, but it is not required as part of this exercise.  Running a pooled logistic regression on data with small time intervals will approximate a Cox proportional hazards model.  Note that using smaller time intervals would be best but this is the data you have (data is not always in the format you wish it was).

```{r}

pooled_model <- glm(div ~ heduc + heblack + mixed + yearscat + heduc:yearscat,
                    data = divorce_EPI608,
                    family = binomial(link = "logit"))

tbl_regression(pooled_model,exponentiate = TRUE)%>%
  as_kable_extra(booktabs = TRUE, longtable = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

### Step 2 	Double your dataset so that it includes counterfactual exposures (education levels) for each couple.

```{r}


data_heduc0 <- divorce_EPI608
data_heduc0$heduc <- 0  

data_heduc1 <- divorce_EPI608
data_heduc1$heduc <- 1  

counterfactual_data <- rbind(data_heduc0, data_heduc1)

counterfactual_data$counterfactual <- rep(c(0, 1), each = nrow(divorce_EPI608))

nrow(divorce_EPI608)
nrow(counterfactual_data)


```

### Step 3 Compute predicted probabilities of remaining married (of “surviving”) for each couple at each time interval and for both counterfactual exposures (education levels).

```{r}

counterfactual_data$pred_div <- predict(pooled_model, 
                                        newdata = counterfactual_data, 
                                        type = "response")

counterfactual_data$pred_survive <- 1 - counterfactual_data$pred_div




aggregate(pred_survive ~ heduc + yearscat, 
          data = counterfactual_data[counterfactual_data$yearscat %in% c(0, 10, 20), ], 
          FUN = mean)
```

### Step 4 Now that you have conditional survival probabilities, cumulatively multiply these probabilities for each couple and counterfactual exposure (education level) to obtain cumulative survival curve data (hint: remember that the sum of log-transformed values is equivalent to a multiplication on the natural scale) Use loops to fill in the data.

```{r}

counterfactual_data <- counterfactual_data %>%
  arrange(id, counterfactual, yearscat) %>%
  group_by(id, counterfactual) %>%
  mutate(cum_survive = cumprod(pred_survive)) %>%
  ungroup()


```

### Step 5 Average cumulative survival probabilities over each year for each counterfactual exposure

```{r}

marginal_curves <- counterfactual_data %>%
  group_by(yearscat, heduc) %>%
  summarise(avg_cum_survive = mean(cum_survive),
            .groups = "drop")


```

### Step 6  Show separate survival curves for each education level category on the same graph.  Label the axes properly. Copy your graph below.

```{r}

survival_plot <- ggplot(marginal_curves, aes(x = yearscat, y = avg_cum_survive, 
                                              color = factor(heduc), 
                                              group = factor(heduc))) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("red", "blue"),
                     labels = c("< 12 Years", "≥ 12 Years"),
                     name = "Husband's Education") +
  labs(x = "Years of Marriage",
       y = "Probability of Remaining Married",
       title = "Marginal Survival Curves: Time to Divorce by Education Level") +
  theme_minimal() +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5))
print(survival_plot)


```

### Question 1

```{r}
marginal_curves %>%
  filter(yearscat %in% c(10, 20, 30, 40)) %>%
  pivot_wider(names_from = heduc, values_from = avg_cum_survive, names_prefix = "heduc_")
```

Education (≥12 years) appears to be detrimental to marriage in the early years but the effect diminishes over time. The blue line (≥12 years education) is consistently below the red line (\<12 years), indicating lower probability of remaining married. However, the curves converge after about 20-30 years, suggesting the negative effect of higher education weakens over time.

### Interpretation Question 2: Difference in 75th percentile of time to divorce

```{r}

marginal_curves %>%
  filter(heduc == 0, avg_cum_survive >= 0.74, avg_cum_survive <= 0.76)

marginal_curves %>%
  filter(heduc == 1, avg_cum_survive >= 0.74, avg_cum_survive <= 0.76)
```

The difference in the 75th percentile of time to divorce between the two education groups is approximately 3 years. Couples where the husband has ≥12 years of education reach the 75th percentile (25% divorced) about 3 years earlier than those with \<12 years of education, indicating faster divorce rates among more educated couples.

### 3 Provided that all usual assumptions are met, how much does education affect the probability of remaining married for 10 years or more? Show your calculation and express your result in words.

```{r}
marginal_curves %>%
  filter(yearscat == 10)
```

Higher education (≥12 years) decreases the probability of remaining married for 10 years or more by 2 percentage points (from 82.4% to 80.4%). This suggests that higher education is associated with a slightly higher risk of divorce within the first 10 years of marriage.

### 4 The survival curves that you generated are marginal. Can you explain why that is? In other words, which of the above step(s) allowed you to weigh estimates according to the observed distribution of covariates? Explain why.

The survival curves are marginal because of Step 5 (averaging cumulative survival probabilities across all individuals). This step averages the predicted survival probabilities over the entire study population for each time point and education level. By doing this, we are standardizing (or weighting) the estimates according to the observed distribution of all other covariates (heblack, mixed) in the actual study population. In other words, the marginal survival curves represent what would happen if we could set everyone's education to a specific level while maintaining their actual values for race and mixed-race status. This creates population-averaged estimates that account for the covariate distribution, rather than conditional estimates for specific covariate combinations.
