---
title: "Assignment 1"
author: "Arkaprabha Gun"
date: "2026-01-27"
output: html_document
---

```{r setup, include=FALSE}
if (!require("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

pacman::p_load(ggdag, dagitty, ggplot2, tidyverse, wesanderson,
               epiR, broom, EValue, stdReg, gtsummary, broom.helpers,epiR,haven,survival,survminer)

set.seed(123)
```


```{r}
addicts2 <- read_dta("data/addicts2.dta")

summary(addicts2)

hist(addicts2$id,breaks = unique(addicts2$id),freq = TRUE)
```

## Question 1 a

```{r}
# Check which records have clinic = 9 or 22
addicts2[addicts2$clinic %in% c(9, 22), ]
addicts2 <- addicts2[addicts2$clinic %in% c(1, 2) & addicts2$survt > 0 & addicts2$survt < 2000, ]


# Create a cleaned dataset excluding these problematic records
addicts_clean <- addicts2[addicts2$clinic %in% c(1, 2), ]

# Verify
table(addicts_clean$clinic)

# Now fit the Kaplan-Meier curves with cleaned data
km_fit <- survfit(Surv(survt, status) ~ clinic, data = addicts_clean)

# Check the fit
km_fit

# Generate the Kaplan-Meier plot with confidence intervals
ggsurvplot(
  km_fit,
  data = addicts_clean,
  conf.int = TRUE,           
  pval = FALSE,              
  risk.table = FALSE,        
  xlab = "Time (days)",
  ylab = "Survival Probability",
  title = "Kaplan-Meier Survival Curves by Clinic",
  legend.title = "Clinic",
  legend.labs = c("Clinic 1", "Clinic 2")
)
```
## Question 1b
```{r}
# Perform the log-rank test
logrank_test <- survdiff(Surv(survt, status) ~ clinic, data = addicts2)

# Display the results
logrank_test
```

The two survival curves are statistically significantly different (p < 0.001). We reject the null hypothesis that the survival curves are the same between the two clinics.
## Question 2
```{r}
plot(km_fit, fun="cumhaz", col=c("red", "blue"), lwd=2,
     xlab="Time (days)", 
     ylab="Cumulative Hazard",
     main="Cumulative Hazard Functions by Clinic")
legend("topleft", c("Clinic 1", "Clinic 2"), col=c("red", "blue"), lty=1, lwd=2)
```
The hazards do not appear to be proportional, Since the curves are not parallel.


## Question 3
```{r}
plot(km_fit, fun="cloglog", col=c("red", "blue"), lwd=2,
     xlab="ln(Time)", 
     ylab="ln(-ln(Survival))",
     main="Complementary Log-Log Plot by Clinic")
legend("topleft", c("Clinic 1", "Clinic 2"), col=c("red", "blue"), lty=1, lwd=2)
```
1. Proportional Hazards (PH) assumption: DOES NOT HOLD

Reasoning: For PH to hold, the lines must be parallel. These lines clearly are not parallel - they converge around ln(time) = 50-100, then diverge again. This confirms what we saw in the cumulative hazard plot.

2. Exponential distribution assumption: DOES NOT HOLD

Reasoning: For an exponential distribution, the lines would need to be straight, parallel, AND horizontal (zero slope). These lines are curved, not parallel, and have positive slopes.

3. Weibull distribution assumption: DOES NOT HOLD

Reasoning: For a Weibull distribution, the lines should be straight and parallel (though they can have a slope). These lines show curvature and are not parallel.

4. Accelerated Failure Time (AFT) assumption: DOES NOT HOLD

Reasoning: AFT models require parallel lines in this plot (similar to PH). Since the lines are not parallel, the AFT assumption does not hold either.


## Question 4
```{r}
# Fit Cox proportional hazards model
cox_model <- coxph(Surv(survt, status) ~ clinic + prison + dose, data = addicts2)

tbl_regression(cox_model, 
               exponentiate = TRUE,
               label = list(
                 clinic ~ "Clinic (2 vs 1)",
                 prison ~ "Prison Record (Yes vs No)",
                 dose ~ "Methadone Dose (per mg/day)"
               ))
```

```{r}
# Test proportional hazards assumption
ph_test <- cox.zph(cox_model)

# Display the results
print(ph_test)

# Plot the Schoenfeld residuals
plot(ph_test)
```
The proportional hazards assumption does NOT hold for the overall model (p = 0.005). Specifically, the assumption is violated for the clinic variable (p < 0.001), while it holds for prison and dose.

```{r}
# Fit Cox model with time-varying clinic effect
cox_tv <- coxph(Surv(survt, status) ~ prison + dose + tt(clinic), 
                data = addicts2,
                tt = function(x, t, ...) x * log(t))


tbl_regression(cox_tv, 
               exponentiate = TRUE,
               label = list(
                 prison ~ "Prison Record (Yes vs No)",
                 dose ~ "Methadone Dose (per mg/day)",
                 `tt(clinic)` ~ "Clinic × log(time)"
               ))
# Calculate clinic HR at median follow-up time
median_time <- median(addicts2$survt)
median_time

# Clinic HR at median time
beta_clinic_time <- coef(cox_tv)["tt(clinic)"]
clinic_hr_median <- exp(beta_clinic_time * log(median_time))
clinic_hr_median

# Calculate 95% CI for clinic at median time
se_clinic <- summary(cox_tv)$coefficients["tt(clinic)", "se(coef)"]
clinic_lower <- exp((beta_clinic_time - 1.96*se_clinic) * log(median_time))
clinic_upper <- exp((beta_clinic_time + 1.96*se_clinic) * log(median_time))

# Create results data frame
results <- data.frame(
  Variable = c("Clinic (2 vs 1, at median time)", "Prison (Yes vs No)", "Dose (per mg/day)"),
  HR = c(clinic_hr_median, exp(coef(cox_tv)["prison"]), exp(coef(cox_tv)["dose"])),
  Lower_CI = c(clinic_lower, exp(confint(cox_tv)["prison",1]), exp(confint(cox_tv)["dose",1])),
  Upper_CI = c(clinic_upper, exp(confint(cox_tv)["prison",2]), exp(confint(cox_tv)["dose",2]))
)

results
```

Clinic 2 performs significantly better than Clinic 1. At the median follow-up time (337 days), patients in Clinic 2 have 68% lower hazard of dropping out compared to Clinic 1 (HR = 0.321, 95% CI: 0.205-0.502, p < 0.001). This protective effect of Clinic 2 strengthens over time.
Patients with a prison record have 44% higher hazard of dropping out (HR = 1.437, 95% CI: 1.033-2.000, p = 0.031). Each additional mg/day of methadone dose reduces the dropout hazard by 3.5% (HR = 0.965, 95% CI: 0.953-0.978, p < 0.001).

## Question 5


```{r}
# Load the flexsurv package
p_load(flexsurv)

# Fit exponential model
exp_model <- flexsurvreg(Surv(survt, status) ~ clinic + prison + dose, 
                         data = addicts2, 
                         dist = "exponential")

# Display summary
summary(exp_model)

# Get coefficients and convert to hazard ratios
# In flexsurvreg with exponential, coefficients are on log-hazard scale
# So HR = exp(coefficient)
exp_results <- data.frame(
  Variable = c("Clinic (2 vs 1)", "Prison (Yes vs No)", "Dose (per mg/day)"),
  Estimate = coef(exp_model)[c("clinic", "prison", "dose")],
  HR = exp(coef(exp_model)[c("clinic", "prison", "dose")])
)

# Get confidence intervals
ci <- confint(exp_model)
exp_results$Lower_CI <- exp(ci[c("clinic", "prison", "dose"), 1])
exp_results$Upper_CI <- exp(ci[c("clinic", "prison", "dose"), 2])

exp_results
```

Clinic 2 performs significantly better than Clinic 1. Patients in Clinic 2 have 58% lower hazard of dropping out compared to Clinic 1 (HR = 0.420, 95% CI: 0.278-0.635). This indicates substantially better patient retention in Clinic 2.
Patients with a prison record have 30% higher hazard of dropping out, though this is not statistically significant (HR = 1.305, 95% CI: 0.943-1.805, CI includes 1.0). Each additional mg/day of methadone dose reduces the dropout hazard by 2.8% (HR = 0.972, 95% CI: 0.960-0.983).

## 5b
```{r}
plot(exp_model, type="hazard", ci=TRUE)

```



The evidence does NOT strongly support the exponential assumption. The complementary log-log plots (Question 3) showed non-linear patterns and non-parallel curves, indicating the hazard is not constant over time. Additionally, the cumulative hazard functions diverged rather than remaining linear and parallel as required by the exponential distribution.


## Question 6

```{r}
# Fit Weibull AFT model
weibull_aft <- flexsurvreg(Surv(survt, status) ~ clinic + prison + dose, 
                           data = addicts2, 
                           dist = "weibull")

# Use tidy to get results
tbl_regression(weibull_aft, 
               exponentiate = TRUE,
               label = list(
                 clinic ~ "Clinic (2 vs 1)",
                 prison ~ "Prison Record (Yes vs No)",
                 dose ~ "Methadone Dose (per mg/day)"
               ))
```
## 6b
```{r}
plot(weibull_aft, type="hazard", ci=TRUE)

```
The Weibull model has a shape parameter = 1.38 . When shape > 1, the hazard increases over time. The exponential is actually a special case of Weibull where shape = 1 (constant hazard). Since our data has shape = 1.38, the hazard increases over time, making the Weibull a better fit than the exponential.


## 6c
```{r}
# Look at the shape parameter and its confidence interval
weibull_tidy[weibull_tidy$term == "shape", ]


AIC(exp_model)
AIC(weibull_aft)
```


 Yes, the baseline hazard is significantly better approximated by the Weibull than the exponential distribution. The shape parameter confidence interval excludes 1.0, and the Weibull model has substantially better fit (lower AIC).
 
 ## 6d

```{r}
# Fit Weibull in proportional hazards form
weibull_ph <- flexsurvreg(Surv(survt, status) ~ clinic + prison + dose, 
                          data = addicts2, 
                          dist = "weibullPH")

# Get results
tidy(weibull_ph, exponentiate = TRUE, conf.int = TRUE)

# Compare coefficients
library(broom)
aft_coef <- tidy(weibull_aft)[c(3:5), c("term", "estimate")]
ph_coef <- tidy(weibull_ph)[c(3:5), c("term", "estimate")]

comparison <- merge(aft_coef, ph_coef, by = "term", suffixes = c("_AFT", "_PH"))
comparison
```

The signs reverse because AFT and PH models parameterize effects differently:
AFT (Accelerated Failure Time):

Models the effect on survival time
Positive coefficient → longer survival time (slower failure)
Example: clinic coefficient = +0.698 means Clinic 2 patients survive longer

PH (Proportional Hazards):

Models the effect on hazard rate (risk of failure)
Positive coefficient → higher hazard (faster failure)
Example: clinic coefficient = -0.961 means Clinic 2 patients have lower hazard

The relationship: Longer survival time ↔ Lower hazard rate
So a positive AFT coefficient (longer survival) corresponds to a negative PH coefficient (lower hazard), and vice versa. They're expressing the same relationship from opposite perspectives.

# Section II Survival Curves
```{r}
 library(haven)
divorce_EPI608 <- read_dta("data/divorce_EPI608.dta")
 View(divorce_EPI608)

str(divorce_EPI608)
head(divorce_EPI608, 20)
summary(divorce_EPI608)

length(unique(divorce_EPI608$id))

table(divorce_EPI608$yearscat)

table(divorce_EPI608$div)

table(divorce_EPI608$heduc)
```

```{r}

pooled_model <- glm(div ~ heduc + heblack + mixed + yearscat + heduc:yearscat,
                    data = divorce_EPI608,
                    family = binomial(link = "logit"))

# Display the model summary
summary(pooled_model)

# Get tidy results
library(broom)
tidy(pooled_model)
```

Education has a time-varying effect (interaction term p = 0.052)
Time decreases divorce risk (negative coefficient)
Mixed race increases divorce risk

Question 2

```{r}
# Question 2: Create counterfactual dataset
# Double the dataset so each couple appears twice:
# Once with heduc = 0, once with heduc = 1

# Create two copies of the data
data_heduc0 <- divorce_EPI608
data_heduc0$heduc <- 0  # Set everyone to <12 years education

data_heduc1 <- divorce_EPI608
data_heduc1$heduc <- 1  # Set everyone to ≥12 years education

# Combine both datasets
counterfactual_data <- rbind(data_heduc0, data_heduc1)

# Add an identifier for which counterfactual exposure
counterfactual_data$counterfactual <- rep(c(0, 1), each = nrow(divorce_EPI608))

# Check the size
nrow(divorce_EPI608)
nrow(counterfactual_data)

# Verify the structure
head(counterfactual_data)
tail(counterfactual_data)
```
## Question 3

```{r}
# Question 3: Compute predicted probabilities of remaining married
# Get predicted probability of divorce for each row
counterfactual_data$pred_div <- predict(pooled_model, 
                                        newdata = counterfactual_data, 
                                        type = "response")

# Convert to probability of remaining married (surviving)
counterfactual_data$pred_survive <- 1 - counterfactual_data$pred_div

# Check the predictions
head(counterfactual_data[, c("id", "yearscat", "heduc", "counterfactual", 
                             "pred_div", "pred_survive")])

# Summary statistics
summary(counterfactual_data$pred_survive)

# Compare by education level
aggregate(pred_survive ~ heduc + yearscat, 
          data = counterfactual_data[counterfactual_data$yearscat %in% c(0, 10, 20), ], 
          FUN = mean)
```

## Question 4
```{r}
# Question 4: Cumulatively multiply survival probabilities (simpler approach)
library(dplyr)

# Calculate cumulative survival using grouped operations
counterfactual_data <- counterfactual_data %>%
  arrange(id, counterfactual, yearscat) %>%
  group_by(id, counterfactual) %>%
  mutate(cum_survive = cumprod(pred_survive)) %>%
  ungroup()

# Check the results
head(counterfactual_data[, c("id", "yearscat", "heduc", "counterfactual", 
                             "pred_survive", "cum_survive")], 20)

# Verify at different time points
counterfactual_data %>%
  filter(yearscat %in% c(0, 5, 10, 15, 20)) %>%
  select(id, yearscat, heduc, counterfactual, cum_survive) %>%
  head(20)
```

# Question 5: Average cumulative survival probabilities by year and education level

```{r}
# This creates the marginal survival curves

marginal_curves <- counterfactual_data %>%
  group_by(yearscat, heduc) %>%
  summarise(avg_cum_survive = mean(cum_survive),
            .groups = "drop")

# View the results
head(marginal_curves, 20)

# Check at specific time points
marginal_curves %>%
  filter(yearscat %in% c(0, 5, 10, 15, 20, 25, 30))
```

## Question 6

```{r}
# Question 6: Plot survival curves by education level
library(ggplot2)

# Create the plot
survival_plot <- ggplot(marginal_curves, aes(x = yearscat, y = avg_cum_survive, 
                                              color = factor(heduc), 
                                              group = factor(heduc))) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("red", "blue"),
                     labels = c("< 12 Years", "≥ 12 Years"),
                     name = "Husband's Education") +
  labs(x = "Years of Marriage",
       y = "Probability of Remaining Married",
       title = "Marginal Survival Curves: Time to Divorce by Education Level") +
  theme_minimal() +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1))

# Display the plot
print(survival_plot)

# Save the plot if needed
# ggsave("survival_curves.png", survival_plot, width = 8, height = 6)
```

## Interp 1
```{r}
# Compare survival probabilities at key time points
marginal_curves %>%
  filter(yearscat %in% c(10, 20, 30, 40)) %>%
  pivot_wider(names_from = heduc, values_from = avg_cum_survive, names_prefix = "heduc_")
```
Education (≥12 years) appears to be detrimental to marriage in the early years but the effect diminishes over time. The blue line (≥12 years education) is consistently below the red line (<12 years), indicating lower probability of remaining married. However, the curves converge after about 20-30 years, suggesting the negative effect of higher education weakens over time.


## Interpretation Question 2: Difference in 75th percentile of time to divorce

```{r}
# Find exact crossing point for 75% survival (0.75)
# For heduc = 0 (< 12 years)
marginal_curves %>%
  filter(heduc == 0, avg_cum_survive >= 0.74, avg_cum_survive <= 0.76)

# For heduc = 1 (≥ 12 years)  
marginal_curves %>%
  filter(heduc == 1, avg_cum_survive >= 0.74, avg_cum_survive <= 0.76)
```
The difference in the 75th percentile of time to divorce between the two education groups is approximately 6 years. Couples where the husband has ≥12 years of education reach the 75th percentile (25% divorced) about 6 years earlier than those with <12 years of education, indicating faster divorce rates among more educated couples.

## Interpretation Question 3: Effect on probability of remaining married for 10+ years

```{r}
# Get survival probabilities at year 10
marginal_curves %>%
  filter(yearscat == 10)
```
Higher education (≥12 years) decreases the probability of remaining married for 10 years or more by 2 percentage points (from 82.4% to 80.4%). In relative terms, couples with higher-educated husbands have 97.6% of the 10-year marriage survival probability compared to those with less-educated husbands. This suggests that higher education is associated with a slightly higher risk of divorce within the first 10 years of marriage.

## Interpretation Question 4: Why are these survival curves marginal?

The survival curves are marginal because of Step 5 (averaging cumulative survival probabilities across all individuals).
This step averages the predicted survival probabilities over the entire study population for each time point and education level. By doing this, we are standardizing (or weighting) the estimates according to the observed distribution of all other covariates (heblack, mixed) in the actual study population.
In other words, the marginal survival curves represent what would happen if we could set everyone's education to a specific level while maintaining their actual values for race and mixed-race status. This creates population-averaged estimates that account for the covariate distribution, rather than conditional estimates for specific covariate combinations.
